\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-09-21 pkg/evtree/inst/doc/evtree.Rnw
started with Section 2, more work to do
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@246 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nojss]{jss}
\usepackage{amssymb, amsmath, amsthm, booktabs, thumbpdf}

@@

@@

The remainder of this paper is structured as follows: Section~\ref{got}
describes the problem of learning globally optimal decision trees and contrasts
it to the locally optimal top-down heuristic that is utilized by recursive
partitioning algorithms. Section~\ref{desc} introduces the \code{evtree} algorithm
before Section~\ref{impl} addresses implementation details along with an
overview of the implemented functions. A benchmark comparison -- comprising
14~benchmark datasets, 3 real-world datasets, and 3 simulated scenarios -- is carried out
in Section~\ref{comp}, showing that the predictive performance of \code{evtree}
is often significantly better compared to commonly used algorithms \code{rpart}
and \code{ctree} (from \pkg{party} package). Finally, Section~\ref{conc} gives
concluding remarks about the implementation and the performance of the new
algorithm.

@@

The aim of classification and regression tree analysis can be stated by
explaining a response variable $Y$ by a vector of $P$ predictor variables
$X=(X_1,...,X_P)$. For classification trees the output variable $Y$ is
qualitative and for regression trees $Y$ is quantitative. Tree-based methods
first partition the input space $X$ into a set of $M$ rectangular regions $R_m$
where $m = 1,...,M$. In the second step a model is fit to each region. The model
fitted to the distribution of $\{Y \in R_m\}$ can be any proper measure of the
location, e.g. the median, or dispersion, e.g. the variance, of a random
variable. Typically the mode is used for classification trees and the arithmetic
mean is applied for regression trees.

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. Then the overall parameter space $\Theta$ can be written
as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the
parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic
element of $\Theta$, then $\theta$ is fully defined by:
%
\begin{equation}
	 \theta=(v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ).
	\label{treeRep}
\end{equation}
%
The overall goal of classification and regression tree analysis can be defined
as finding a set of parameters $\hat{\theta}$, such that
%
\begin{equation}
	\hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; loss(Y, f(X,\theta)) + comp(\theta).
	\label{generalcostfunction}
\end{equation}
%
Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives
predictions for the output $Y$. $loss$ is a suitable loss function for the
domain of $Y$. Commonly, the misclassification error for classification
problems, or the squared error for regression problems, is used. $comp$
quantifies the complexity of trees and penalizes for more complex models in the
tree selection process. In CART, the complexity is a non-negative function of
the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all
$\Theta_m$.

@@

Instead of searching all combination of $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding
split point $s_r$ is selected such that it minimizes the impurity in the two
resulting terminal nodes. Let the parameters of a tree, constructed in
$i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} =
(\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the
position of the new internal node $r$, as well as the corresponding split
variable $v_r$ and split point $s_r$, are chosen such that:
%
\begin{equation}
		\min_{v_r, s_r} \; loss(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
	\label{eq:whichsplit}
\end{equation}
%
CART's standard measures for node impurity are the Gini index, for
classification problems and the mean squared error, for regression problems.
After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by
$\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This
process of recursively selecting and executing splits continues until the
homogeneity of the nodes cannot be further improved. Additionally, often
additional stopping criteria are defined, e.g. minimum sample sizes in the
internal nodes and terminal nodes.

In CART, and many other recursive tree methods, the tree growing step is
followed by a pruning step. The idea behind pruning is to first grow a large
tree without or with little restriction on the magnitude of impurity reduction.
Starting from the bottom of the tree pruning then removes worthless splits. The
method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and
used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning
iteratively prunes the internal node that:
%
\begin{equation}
		\min_{r=1,...,n_{M-1}} \; loss(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + comp(\tilde{\theta}_i \setminus (v_r, s_r)).
	\label{eq:pruning}
\end{equation}
%
The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus
(\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.

@@

\begin{equation}
	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; 2 \cdot MC(Y, f(X, \theta))+\alpha \cdot M \cdot \ln N.
\label{eq:classification}
\end{equation}

@@

\begin{equation}
	 \hat{\theta} = \operatorname*{\arg \,\min}_{\theta \in \Theta} \; N \cdot \ln \mathit{MSE}(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
\label{eq:regression}
\end{equation}

>>>>

\documentclass[nojss]{jss}
\usepackage{amssymb, amsmath, amsthm, booktabs, thumbpdf}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

@@

\pagebreak

@@

The remainder of this paper is structured as follows: Section~\ref{got}
describes the problem of learning globally optimal decision trees and contrasts
it to the locally optimal forward-search heuristic that is utilized by recursive
partitioning algorithms. Section~\ref{desc} introduces the \code{evtree} algorithm
before Section~\ref{impl} addresses implementation details along with an
overview of the implemented functions. A benchmark comparison -- comprising
14~benchmark datasets, 3~real-world datasets, and 3~simulated scenarios -- is carried out
in Section~\ref{comp}, showing that the predictive performance of \code{evtree}
is often significantly better compared to the commonly used algorithms \code{rpart}
and \code{ctree} (from the \pkg{party} package). Finally, Section~\ref{conc} gives
concluding remarks about the implementation and the performance of the new
algorithm.

@@

Classification and regression tree analysis aims at
modeling a response variable $Y$ by a vector of $P$ predictor variables
$X=(X_1,...,X_P)$. For classification trees the output variable $Y$ is
qualitative and for regression trees $Y$ is quantitative. Tree-based methods
first partition the input space $X$ into a set of $M$ rectangular regions $R_m$
($m = 1,...,M$) then fit a (typically simple) model within each region $\{Y | X \in R_m\}$,
e.g., the mean, median, or variance etc. Typically, the mode is used for
classification trees and the arithmetic mean is applied for regression trees.

@@

As done by \cite{breiman1984cart}, let the complexity of a tree be measured by a
function of the number of terminal nodes, without further considering the depth
or the shape of trees. Then the overall parameter space $\Theta$ can be written
as $\Theta =\bigcup_{M=1}^{M_{max}} \Theta_{M}$, where $\Theta_M$ denotes the
parameter space of models with $M$ terminal nodes. Let $\theta$ be a generic
element of $\Theta$, then $\theta$ is fully defined by:
%
\begin{equation}
	 \theta=(v_{n_1}, s_{n_1},...,v_{n_{M-1}},s_{n_{M-1}} ).
	\label{treeRep}
\end{equation}
%
The overall goal of classification and regression tree analysis can be defined
as finding a set of parameters $\hat{\theta}$, such that
%
\begin{equation}
	\hat{\theta} = \argmin_{\theta \in \Theta} \; \mathit{loss}(Y, f(X,\theta)) + \mathit{comp}(\theta).
	\label{generalcostfunction}
\end{equation}
%
Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives
predictions for the output $Y$. $loss$ is a suitable loss function for the
domain of $Y$. Commonly, the misclassification error for classification
problems, or the squared error for regression problems, is used. $comp$
quantifies the complexity of trees and penalizes for more complex models in the
tree selection process. In CART, the complexity is a non-negative function of
the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all
$\Theta_m$.

@@

Instead of searching all combination of $\Theta$ simultaneously, recursive
partitioning algorithms only consider one split at a time. At each internal node
$r \in \{n_1,..., n_{M-1}\}$, the split variable $X_{v_r}$ and the corresponding
split point $s_r$ is selected such that it minimizes the impurity in the two
resulting terminal nodes. Let the parameters of a tree, constructed in
$i$-steps, be denoted as $\tilde{\theta}_{i}$ and $\tilde{\theta}_{0} =
(\emptyset)$. Then, for a tree with parameters $\tilde{\theta}_{i}$, the
position of the new internal node $r$, as well as the corresponding split
variable $v_r$ and split point $s_r$, are chosen such that:
%
\begin{equation}
		\min_{v_r, s_r} \; \mathit{loss}(Y, f(X, \tilde{\theta}_{i-1} \cup (v_r, s_r) ))
	\label{eq:whichsplit}
\end{equation}
%
CART's standard measures for node impurity are the Gini index, for
classification problems and the mean squared error, for regression problems.
After the selection of the $i^{th}$ split $\tilde{\theta}_{i}$ is given by
$\tilde{\theta}_{i} = \tilde{\theta}_{i-1}  \cup (\hat{v}_r, \hat{s}_r)$. This
process of recursively selecting and executing splits continues until the
homogeneity of the nodes cannot be further improved. Additionally, often
additional stopping criteria are defined, e.g. minimum sample sizes in the
internal nodes and terminal nodes.

In CART, and many other recursive tree methods, the tree growing step is
followed by a pruning step. The idea behind pruning is to first grow a large
tree without or with little restriction on the magnitude of impurity reduction.
Starting from the bottom of the tree pruning then removes worthless splits. The
method proposed by Breiman \textit{et al.} \cite[Chapter 2]{breiman1984cart} and
used by CART is called \textit{cost-complexity pruning}. Cost-complexity pruning
iteratively prunes the internal node that:
%
\begin{equation}
		\min_{r=1,...,n_{M-1}} \; \mathit{loss}(Y, f(X,\tilde{\theta}_i \setminus (v_r, s_r )))  + \mathit{comp}(\tilde{\theta}_i \setminus (v_r, s_r)).
	\label{eq:pruning}
\end{equation}
%
The subsequent solution $\tilde{\theta}_{i+j+1} = \tilde{\theta}_{i+j} \setminus
(\hat{v}_r, \hat{s}_r)$, where $j$ is the number of completed pruning steps.

@@

\begin{equation}
	 \hat{\theta} = \argmin_{\theta \in \Theta} \; 2 \cdot MC(Y, f(X, \theta))+\alpha \cdot M \cdot \ln N.
\label{eq:classification}
\end{equation}

@@

\begin{equation}
	 \hat{\theta} = \argmin_{\theta \in \Theta} \; N \cdot \ln \mathit{MSE}(Y, f(X, \theta))+\alpha \cdot 4 \cdot (M+1) \cdot \ln N.
\label{eq:regression}
\end{equation}
